A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-28 paper/writeup.tex
added missing caption
peter edits to the edits

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    % NSP-22NOV: "unary node" and "ARG" are not defined so I tried to provide better context here
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees where they
    are not coalescence events.
    We show how these non-coalescing haplotypes (``unary regions'') may be
    inserted into the ancestral recombination graph (ARG), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting inferred ARGs are smaller, faster to compute with,
    and contain substantially more information about inheritance % NSP-22NOV: "substantially more information" in what regard?
    than inferred ARGs without unary nodes.
    We provide efficient algorithms to infer unary nodes within existing ARGs,
    new metrics of agreement and disagreement between inferred ARGs,
    and explore some consequences of unary nodes for ARGs inferred from real data.
\end{abstract}

@@

Haplotype identity has been largely overlooked in the literature on ARG inference so far --
in that the metrics that have been used to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on the sharing of ancestral haplotypes that span across these trees. % NSP-22NOV: what does "sharing of ancestral haplotype identities" mean, concretely? Would "... depend only on the sequence of marginal trees, but not on the ancestral haplotypes that span across these trees" be a clearer description?
% HF-24NOV: Noted and changed
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describe the relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples % NSP-22NOV: the meaning here isn't clear to me-- what does "represent the focal set of genotyped samples" mean, concretely?
% HF-24NOV : I am not exactly sure, I will ping Peter
can be unambiguously said to represent the same thing in both.
Thus, metrics that only describe these relationships
ignore the additional ancestral information carried by the structure of other haplotypes. % NSP-22NOV: again, "additional information carried by the structure of other haplotypes" is vague
% HF-24NOV : Heard
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure --
two ARGs could be identical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure the sharing of haplotypes between adjacent trees.

@@

\algorithmref{alg:extend} simplifies the algorithm in several ways for the sake of clarity --
for instance, the bookeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made.
The main step that is omitted is a description of the \texttt{merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, see \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

@@

It is helpful to first describe what we compute in the simple case.
\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
for each node the total span in a ``true'' tree sequence, 
given by the line,
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
\citep{kelleher2018efficient},
which removes all unary nodes from all trees
thus reducing the node spans for some nodes in the sequence.
In \figref[D]{fig:node-spans}
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
\figref[B]{fig:node-spans} shows how much of this span is incorrect:
for each node, it gives the amount of span
that is incorrect (i.e., not present in the original tree sequence) % NSP-22NOV: it seems a bit odd that this panel (B) is not on a log scale while all the others are
% HF-24NOV Peter and I looked at both options and we thought the log scale on the Y-axis looked a bit misleading against our method. We instead provide our suplemental figure to provide more information about Figure 5b
over the total amount of span added after running \algorithmref{alg:extend}.
The values in these plots
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how the node spans represented in the result compare to those in the original tree sequence. % NSP-22NOV: "relationships" of what exactly?
% HF -24NOV: Noted and changed
Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are otherwise unclear as to the equivalency of nodes across sequences
(as for instance with a true, simulated tree sequence and one inferred from its genotypes;
in the former nodes represent actual ancestral haplotypes, and in the latter
represent hypothetical ancestors which may or may not resemble the truth).
Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
and the same set of sample nodes.
We would like to measure (a) how much of $T_1$ is found in $T_2$;
(b) how much of $T_1$ is \emph{not} found in $T_2$; and
(c) how much of $T_2$ is not found in $T_1$.
(Think of these three quantities as the size of the intersection
and two differences between the tree sequences,
thought of vaguely as sets.)
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(\T_1, n1; \T_2, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the subset of all places $k$ where $n_1$ and $n_2$ share the same sample set in $T_k^{(1)}$ and $T_k^{(2)}$, respectively.
Further, we define
% NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
% $\mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]$
% and also in next equation for consistency.
% HF24-NOV: I think I disagree but we can get a third opinion.
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(\T_1, n1; \T_2, n2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
% NSP-22NOV: as before I'd suggest using \mathbb{I}[...] for the indicator function
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k,\ n_2) = s(T^{(1)}_k,\ n)} 
\end{align*}
as the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could possibly be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

The implementation of this method in \texttt{tscompare}
additionally produces relative values:
the ancestral recombination graph Robinson-Foulds value (relative dissimilarity) and true proportion represented. 
The \emph{ARG Robinson-Foulds} (ARF) is defined to be the dissimilarity proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert}.
\end{align}
The \emph{true proportion represented} (TPR) is similarity between two trees relative 
to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
	{\Vert \T_2\Vert}.
\end{align}
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
We compute dissimilarity and ARF of two tree sequences in \figref{fig:conceptual_discrepancy}.

@@

To do this, we simulate ARGs containing full haplotypes using \msprime's 
option \texttt{coalescing\_segments\_only} set to False.
Although \msprime\ simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime\ are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
More concretely, we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the coalescent segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, when evaluating what proportion of the true ARG is represented in an inferred ARG
we omit these isolated non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

In the simple example in \figref{fig:conceptual},
extending haplotypes replaces three edges
($0 \to 3$ and $3 \to 4$ on the left tree, and $0 \to 4$ on the right tree)
by two edges ($0 \to 3$ and $3 \to 4$ on both trees).
If all edge endpoints were unique, then we'd expect \emph{every} edge to be extendable
on one of its ends
(except those pendant to the root and some of those adjacent to chromosome ends),
leading to a reduction in number of edges by a third.
Experiments with an earlier version of the algorithm showed that
if we only extend haplotypes on ``paths of length 1'', 
then the hypothesized reduction of 1/3 is achieved for long sequences. % NSP-22NOV: is this sentence mentioning the earlier algorithm necessary?
% HF-24NOV: No, but I kind of like the homage to our own earlier work and it sort of makes a benchmark for edge reduction proportions.
It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
but we still expect the number of edges to decrease by more than 1/3.
Indeed,
\figref{fig:speed_and_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
in half, as long as the sequence is long enough.

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
Indeed, \tsinfer{} provides inferred ARGS already containing unary nodes;
to what degree are they taking advantage of this information already?
Here, we study the behavior of \tsinfer{}, which produces unary nodes as a
byproduct of its inference algorithm,
and leave a comparison across other ARG inference methods for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
% NSP-22NOV: do we need to define TS simplification here (or earlier?)
% HF-24NOV: We define in the previous section line 960.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 20\% (bottom panels),
but that roughly 10\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span.
Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.)

@@

\begin{figure}
    \includegraphics{benchmarks/constant_pop_results_ratios}
    \caption{
        As in \figref{fig:speed_and_edges} except that
        the original tree sequence was simulated with an constant population
        of size $10^4$ (using the ``constant dog'' scenario);
        see Methods for details.
        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_constant_other}.
        \label{fig:speed_and_edges_constant}
    }
\end{figure}

>>>>

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees in which they
    do not represent coalescences.
    We show how these non-coalescing haplotypes (``unary regions'') may be
    inserted into the ancestral recombination graph (ARG), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting ARGs are smaller, faster to compute with,
    and the additional ancestral information that is inserted
    is nearly always correct where the initial ARG is correct.
    We provide efficient algorithms to infer unary regions within existing ARGs,
    and explore some consequences of unary nodes for ARGs inferred from real data.
    To do this, we introduce new metrics of agreement and disagreement between inferred ARGs
    that, unlike previous methods, consider the ARG as describing relationships between haplotypes
    rather than just a collection of trees.
\end{abstract}

@@

Haplotype identity has been largely overlooked in the literature on ARG inference --
to our knowledge, all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on how ancestral haplotypes span across these trees.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure --
two ARGs could be identical by either measure
but have completely different identity by descent (IBD) segment structure.
% (as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure the sharing of haplotypes between adjacent trees.
However, the additional information provided by haplotype structure can be important:
for instance, haplotypes that extend over many marginal genealogies
``tie together'' those genealogies, allowing estimates of times of particular ancestors
to be informed by larger portions of the genome on which there are many genealogies.

@@

\algorithmref{alg:extend} simplifies the algorithm in several ways for the sake of clarity --
for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made.
The main step that is omitted is a description of the \texttt{merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, see \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

@@

It is helpful to first describe what we compute in the simple case.
\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
for each node the total span in a ``true'' tree sequence, 
given by the line,
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
\citep{kelleher2018efficient},
which removes all unary nodes from all trees
thus reducing the node spans for some nodes in the sequence.
In \figref[D]{fig:node-spans}
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
\figref[B]{fig:node-spans} shows how much of this span is incorrect:
for each node, it gives the amount of span
that is incorrect (i.e., not present in the original tree sequence)
over the total amount of span added after running \algorithmref{alg:extend}.
The values in these plots
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how the node spans represented in the result compare to those in the original tree sequence.
Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are otherwise unclear as to the equivalency of nodes across sequences
(as for instance with a true, simulated tree sequence and one inferred from its genotypes;
in the former nodes represent actual ancestral haplotypes, and in the latter
represent hypothetical ancestors which may or may not resemble the truth).
Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
and the same set of sample nodes.
We would like to measure (a) how much of $T_1$ is found in $T_2$;
(b) how much of $T_1$ is \emph{not} found in $T_2$; and
(c) how much of $T_2$ is not found in $T_1$.
(Think of these three quantities as the size of the intersection
and two differences between the tree sequences,
thought of vaguely as sets.)
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).

Concretely, our method can be applied to any two tree sequences,
as in the case of \figref{fig:conceptual_discrepancy},
where we consider the pair of trees which exist at that break point.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(\T_1, n1; \T_2, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
Further, we define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(\T_1, n_1; \T_2, n_2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let $s(\T,n)$ denote
the total span that node $n$ is present in the marginal trees,
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,][]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)1}^N (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

The implementation of this method in \texttt{tscompare}
additionally produces relative values:
the ancestral recombination graph Robinson-Foulds value (relative dissimilarity) and true proportion represented. 
The \emph{ARG Robinson-Foulds} (which we call ``ARF'') is defined to be the dissimilarity proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert}.
\end{align}
The \emph{true proportion represented} (TPR) is similarity between two trees relative 
to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
	{\Vert \T_2\Vert}.
\end{align}
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
We compute dissimilarity and ARF of two tree sequences in \figref{fig:conceptual_discrepancy}.

@@

To do this, we simulate ARGs containing full haplotypes using \msprime's 
option \texttt{coalescing\_segments\_only} set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
More concretely, we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, when evaluating what proportion of the true ARG is represented in an inferred ARG
we omit these isolated non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

In the simple example in \figref{fig:conceptual},
extending haplotypes replaces three edges
($0 \to 3$ and $3 \to 4$ on the left tree, and $0 \to 4$ on the right tree)
by two edges ($0 \to 3$ and $3 \to 4$ on both trees).
If all edge endpoints were unique, then we'd expect \emph{every} edge to be extendable
on one of its ends
(except those pendant to the root and some of those adjacent to chromosome ends),
leading to a reduction in number of edges by a third.
Experiments with an earlier version of the algorithm showed that
if we only extend haplotypes on ``paths of length 1'', 
then the hypothesized reduction of 1/3 is achieved for long sequences.
It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
but we still expect the number of edges to decrease by more than 1/3.
Indeed,
\figref{fig:speed_and_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
in half, as long as the sequence is long enough.

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
Indeed, \tsinfer{} provides inferred ARGS already containing unary nodes;
to what degree are they taking advantage of this information already?
Here, we study the behavior of \tsinfer{}, which produces unary nodes as a
byproduct of its inference algorithm,
and leave a comparison across other ARG inference methods for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 20\% (bottom panels),
but that roughly 10\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span.
Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.)

@@

\begin{figure}
    \includegraphics{benchmarks/constant_pop_results_ratios}
    \caption{
        As in \figref{fig:speed_and_edges} except that
        the original tree sequence was simulated with an constant population
        of size $10^4$ (using the ``constant dog'' scenario);
        see Methods for details.
        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_other_constant}.
        \label{fig:speed_and_edges_constant}
    }
\end{figure}
